---
title: "wrangling"
output: html_document
---

```{r}
library(tidyverse)
#library(plotly)
library(skimr)

```

```{r}
gapminder <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv')

gapminder

```





```{r}
head(gapminder,6)
str(gapminder)
summary(gapminder)
skimr::skim(gapminder)

names(gapminder)
nrow(gapminder)
ncol(gapminder)
dim(gapminder)

c(nrow(gapminder),ncol(gapminder)) #same as dim


```


***dplyer**

verbs = 
*filter*
subset by row
subset by column
*mutate* = create new column

*summarize* down
*arrange* (reorder)
```{r}
#filter by rows
filter(gapminder, lifeExp <31)
filter(gapminder, country=="Mexico")

#adding two or more comparators in a text comparision using %in% and c()
filter(gapminder, country %in% c("Mexico", "Peru"))
filter(gapminder, country == "Mexico", year == 2002)
filter(gapminder, country == "Mexico", lifeExp < 50)

#find mean LifeEx of Sweden

sweden.data <- filter(gapminder, country == "Sweden")

# to select a named column of sweden.data us "$"
mean.sweden.data <- mean(sweden.data$lifeExp)


```



```{r}
select(gapminder, year, lifeExp)
select(gapminder, -continent, -lifeExp) # to exclude these columns

#use select and filter together
gap_cambodia <- filter(gapminder, country == "Cambodia")
gap_cambodia2 <- select(gap_cambodia, -continent, -lifeExp)

```

***Pipes***
somewhat different than unix. have to escape operators


```{r}
gapminder %>% head(3)
gap_cambodia3 <- gapminder %>% filter(country == "Cambodia") %>% select(-continent, -lifeExp)

#clearly written: white space doesn't matter here
gap_cambodia3 <- gapminder %>%
  filter(country == "Cambodia") %>%
  select(-continent, -lifeExp)



``` 
***Mutate***

```{r}
gapminder %>% 
  mutate(index = 1:nrow(gapminder))
gapminder %>% 
  mutate(gdp= pop* gdpPercap)

egypt_vietnam <- filter (gapminder, country %in% c("Egypt", "Vietnam"))

gapminder %>%
    filter (country %in% c("Egypt", "Vietnam")) %>%
    mutate (gdp=pop*gdpPercap, max_gdp =max(gdp))


```

**using group_by to analyze a dataset by a particular variable**
group_by(variable)
using mutate gives multiple of the same entry for each same value of the variable

```{r}
gdp_grouped <- gapminder %>% 
  group_by(country) %>% 
  mutate(gdp =pop * gdpPercap, max_gdp=max(gdp)) %>%
  ungroup()

View(gdp_grouped)
  
```

**using summarize to get out a single value per particular group variable**
reduces data frame to grouped varaible and analysis variable


```{r}
gdp_summarized <- gapminder %>% 
  group_by(country) %>% 
  mutate(gdp =pop * gdpPercap, max_gdp=max(gdp)) %>%
  summarize(max_gdp = max(gdp)) %>% 
  ungroup()

View(gdp_summarized)
  
```
**arrange**


```{r}
gdp_summarized_sort <- gapminder %>% 
  group_by(country) %>% 
  mutate(gdp =pop * gdpPercap, max_gdp=max(gdp)) %>%
  summarize(max_gdp = max(gdp)) %>% 
  ungroup() %>% 
  arrange(desc(max_gdp))

View(gdp_summarized_sort)
  
```

***dplyr***
data wrangling by joining/union/intersection of data tables (frames)
adding CO2

```{r}
## read in the data. (same URL as yesterday, with co2.csv instead of gapminder.csv)
co2 <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv")
```


```{r}
## explore
co2 %>% head()
co2 %>% dim() # 12

## create new variable that is only 2007 data
gap_2007 <- gapminder %>%
  filter(year == 2007) 
gap_2007 %>% dim() # 142  



##left join
lj <- left_join(gap_2007, co2, by="country")
lj <- gap_2007 %>% 
  left_join(co2, by ="country")

rj <- right_join(gap_2007, co2, by ="country")

lj2 <-left_join(co2, gap_2007, by = "country")

```

***More data wrangling***

```{r}

## read in gapminder data in wide format
gap_wide <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv')



```
**gather**
##tidyr::gather()

```{r}
gap_long <- gap_wide %>% 
  gather(key= obstype_year,
         value=obs_values,
         -country, -continent)
```

## tydr:separate()

```{r}
gap_long <- gap_wide %>% 
  gather(key= obstype_year,
         value=obs_values,
         -country, -continent) %>% 
  separate(obstype_year,into=c("obs_type","year"),sep="_", convert=TRUE)

```
##plot


```{r}
canada_df <- gap_long %>% 
  filter(country == "Canada",
         obs_type == "lifeExp")

ggplot(data=canada_df, aes(x=year, y=obs_values)) + 
  geom_line()

```



```{r}
americas_df <- gap_long %>% 
  filter(continent == "Americas",
         obs_type == "lifeExp")
ggplot(data=americas_df, aes(x=year, y=obs_values, color=country)) + 
  geom_line()

```

##Exercise


```{r}
continent_df <- gap_long %>% 
  filter (obs_type == "lifeExp", year> 1980) %>% 
  group_by (continent, year) %>%
  summarize (mean_lifeExp = mean(obs_values)) %>%
  ungroup()

ggplot(data=continent_df, aes(x=year, y=mean_lifeExp, color=continent)) +
         geom_line()+
  labs(y ="Age",
       x= "Year",
       title= "Mean life expectancy")
```

## spread()

```{r}
gap_normal <- gap_long %>% 
  spread(key= obs_type, value=obs_values)
  
  
  
```

##Exercise gap long to gap_wide again

```{r}
gap_wide2 <- gap_long %>%
  unite(type_year,obs_type,year, sep="_") %>%
  spread(key=type_year, value=obs_values)


```

